% !TEX program = tectonic
\documentclass[anon]{nesy2026}   % submission: [anon]; camera-ready: [final]

% --- light extras the class does NOT auto-load ---
\usepackage{booktabs}              % nicer tabular rules
\usepackage{makecell}              % multi-line table cells (§2.6 capability matrix)

% --- propositions: JMLR class predefines proposition, theorem, lemma,
% --- remark, example. No \newtheorem needed. Use \begin{proposition}...
% --- \end{proposition} directly with \label{prop:...} for \theoremref.

% --- inline code-span macro ---
\newcommand{\C}[1]{\texttt{#1}}

% --- title + author (JMLR-style) ---
\title[Persistence OS]{Toward Accountable Neurosymbolic Runtimes: The Persistence OS Substrate}

% Author block — class option [anon] auto-suppresses names in submission PDF.
% For camera-ready, flip the documentclass option to [final] and the same author
% block renders with names.
\clearauthor{%
  \Name{Author Name}                       % TODO: real name pre-camera-ready
  \Email{anonymous@example.com}            % TODO: real email pre-camera-ready
  \\\addr Affiliation                      % TODO: real affiliation pre-camera-ready
}

\begin{document}

\maketitle

% --- Revision history ---
% Intentionally NOT included in submission .tex. The v0.1-v0.7 history is
% preserved in paper/archive/persistence-nesy-2026-v0.7.md (lines 10-19).
% If a future reviewer requests it, re-add as a \appendix section in the
% camera-ready (\documentclass[final]{nesy2026}) build only.

\begin{abstract}
We present \textbf{Persistence}, a neurosymbolic substrate where symbolic state --- bitemporal datoms, content-addressed plan ASTs, policy-as-data, and boundary specs --- carries formal properties (immutability, replay determinism, audit-chain Merkle integrity, plan content-addressing with descendant propagation) on top of which neural agency composes. The central design invariant is that every piece of agent state --- memory, audit, plan, skill, transaction --- is an immutable, content-addressed, bitemporal \emph{datom}; effects route through a composable handler stack whose entries are themselves datoms; plans are EDN abstract syntax trees stored as Merkle-DAGs; counterfactual replay is a first-class query over the log. Large-language-model (LLM) agents are moving into regulated, high-stakes domains --- project finance, insurance, algorithmic trading, clinical operations, hospitality --- where auditability, counterfactual analysis, and controllable learning are not optional, and contemporary agent frameworks fragment these requirements across incompatible systems: temporal knowledge graphs for memory~\citep{kurtic2025zep}, algebraic effect systems for tool use~\citep{cheng2025pangolin}, declarative program synthesis for reasoning~\citep{khattab2023dspy}, skill libraries for continual learning~\citep{wang2023voyager,memento2026skills}, and seed-replay kludges for counterfactuals~\citep{anon2026camo,anon2026agenther}. Persistence's claim is that these properties can be \emph{derived} from a single symbolic substrate rather than re-engineered per system. This paper reports \textbf{Phase 1 and the first Phase-2 increment of the reference implementation (v0.1.0a1 + v0.4.0a1, 832 tests green, 7 xfailed)}, which ships five of the seven runtime modules --- Fact, Effect, Spec, Replay, and Plan (minimal: parse / unparse / walk / spec-validation with 128-bit content-addressing) --- and demonstrates five substrate-derived capabilities: queryable bitemporal history, counterfactual branching, composable policy-gated effects, boundary-checked neurosymbolic contracts, and content-addressed homoiconic plan ASTs with byte-identical round-trip. Two further modules (Transactions, REPL) remain specified-but-not-shipped; Plan-module \emph{execution} (effect-dispatched evaluation, Pareto-vector fitness, MIPROv2 / MCTS / evolutionary optimization, 4-gate skill promotion) is Phase 2.C. We formalize the substrate (bitemporal datom model, effect-handler stack with a machine-checkable well-formedness property, counterfactual replay with byte-identical NO-OP invariance, Merkle-hashed audit chain, parse-don't-validate boundary specs with LLM-self-healing hints) and prove five formal propositions that hold on the shipped code; the headline two are (i) any handler stack is well-formed iff every catalog operation is covered --- checkable in linear time by \C{Runtime.is\_well\_formed}; (ii) \C{replay(T, I)} with a NO-OP intervention yields a trajectory whose canonical hash is byte-identical to the factual trajectory's --- a stronger determinism guarantee \textbf{for the NO-OP intervention case} than CAMO's aspirational seed replay (for non-trivial interventions, byte-identity no longer applies; the suffix diverges as soon as the intervened action changes observations). We describe a Reproduction Plan for four benchmarks --- LongMemEval, CAMO-style counterfactual fidelity, a novel 50-trajectory synthetic regulator-replay, and one production case study --- to be populated by the camera-ready. The neurosymbolic positioning is load-bearing: the substrate is explicitly symbolic (bitemporal datom model with a Datalog-shaped query surface, EDN-grammar plan-ASTs, policy-as-data, Malli-style specs) while agency remains neural; the properties this paper proves are properties of the symbolic layer under the neural one.
\end{abstract}

\section{Introduction}\label{sec:intro}

\textbf{Research question.} When neural agency is scaffolded by a symbolic substrate, what formal properties does the combined system inherit that neither layer provides alone? We argue that six substrate invariants --- immutable bitemporal datoms, algebraic-effect handlers with audit semantics, homoiconic plan ASTs, transactional shared state, machine-readable boundary specs, and a REPL --- yield derived capabilities (accountability, replay, counterfactual fidelity, composable safety, compositional skill learning, multi-agent coordination) that are consequences of the substrate, not independently-engineered features. This is the neurosymbolic bet Persistence makes: the \emph{symbolic} layer is designed to carry formal guarantees as proved invariants of the log; the \emph{neural} layer (LLM decisions, embeddings, generative skills) stays expressive and open-ended above it; the interesting properties emerge at the interface.

Large-language-model agents are no longer research demonstrations. In the past eighteen months, agents have reached production in project-finance assessment, insurance quote aggregation, hospitality operations, algorithmic trading, clinical intake, legal drafting, and hundreds of smaller verticals. The production gap has exposed a class of problems the framework literature has not addressed --- and, seen through the NeSy lens, each is a place where the missing ingredient is symbolic-substrate structure under the neural layer:

\begin{enumerate}
\item \textbf{Accountability.} Regulators asking ``what did the agent believe at 14:03 on April 14, and why did it decide X?'' receive hand-waved answers assembled from prose logs. There is no formal substrate for temporal provenance.
\item \textbf{Counterfactual reasoning.} Post-incident analysis --- \emph{``what if the agent had waited ten minutes?''} --- is either absent or simulated by re-running the agent with a perturbed prompt, which changes far more than the intended variable.
\item \textbf{Controllable learning.} Agents that ``improve themselves'' by rewriting markdown skill files or mutating memory in place accumulate semantic drift and procedural drift~\citep{hannecke2026ssgm}.
\item \textbf{Composable safety.} Guardrail frameworks (NeMo Guardrails, LLM Guard, Rebuff) are single-layer interceptors. Multi-tenant, regulated agents require stackable policy, dry-run, cache, rate-limit, and audit layers with well-defined interaction semantics.
\item \textbf{Plan opacity.} Prose chain-of-thought is inspectable but not editable. JSON tool-call graphs are editable but not expressive. Neither is a first-class program the agent can reason over.
\end{enumerate}

The field has converged on partial solutions. Zep's Graphiti introduces bitemporal knowledge graphs~\citep{kurtic2025zep}; Pangolin~\citep{cheng2025pangolin} and~\citet{wang2025composable} formalize algebraic effects for LLM programming; DSPy treats agents as declarative programs~\citep{khattab2023dspy}; Voyager~\citep{wang2023voyager} and Memento-Skills~\citep{memento2026skills} construct executable skill libraries; CAMO~\citep{anon2026camo} and AgentHER~\citep{anon2026agenther} formalize counterfactual replay via aligned randomness. Each is a beachhead. None are composed. The resulting integration burden --- five substrates, five consistency models, five failure modes --- erodes the properties each was designed to provide.

\textbf{Contribution.} We present \textbf{Persistence}, a cognitive-runtime substrate built around a single claim: accountability, replay, counterfactuals, composable safety, and compositional skill learning can be derived from \emph{one} substrate rather than re-engineered per system. The substrate treats every piece of agent state --- memory facts, audit entries, plan-AST nodes, skill-library entries, transaction commits --- as an immutable, content-addressed, bitemporal \emph{datom}. Effects route through a composable handler stack whose entries are themselves datoms. Plans are EDN abstract syntax trees stored as datom graphs; skills are named, content-addressed subtrees. Trajectories are ordered sequences of effect datoms under a shared run-id, making counterfactual replay a first-class query over the log. Transactions compose via software transactional memory over datom-backed refs. Specs constrain every boundary. A REPL module exposes inspection, editing, rewind, and speculative branching against running agents.

\textbf{What this paper reports, honestly.} Phase 1 of the reference runtime (v0.1.0a1, tagged 2026-04-20) shipped four of the seven modules --- Fact, Effect, Spec, Replay --- with 579 passing tests. The first Phase-2 increment ships the fifth module, Plan, in a deliberately minimal form: parse, unparse, depth-first walk, and spec validation over a uniform \C{[tag \{attrs\} *children]} EDN AST, with 128-bit content-addressing via \C{sha256(canonical(node))[:32]}. The Plan module merged at v0.2.0a1 on 2026-04-24 and received three patch releases over the following day: v0.2.0a2 hardened spec-error composition, alias preservation via \C{:original-tag}, and schema evolution; v0.2.0a3 closed a reserved-attr-key round-trip falsifier surfaced by hypothesis at \C{max\_examples=200}; v0.3.0a1 (2026-04-25) shipped a static-plus-manifest coercion registry that lets \C{datetime} / \C{bytes} / \C{Decimal} / \C{UUID} / \C{frozenset} / \C{edn\_format.Symbol} participate in canonical hashing without breaking content-addressing determinism, and introduced a \C{PLAN\_CANONICAL\_VERSION} manifest constant for schema-evolution callers. The combined suite at v0.1.0a1 + v0.4.0a1 runs \textbf{832 tests green, 7 xfailed}. The Plan-module \emph{execution} runtime (effect-dispatched evaluation, optimizer ladder, 4-gate skill promotion) remains Phase 2.C. Together the five shipped modules demonstrate five substrate-derived capabilities as shipped: queryable bitemporal history (\C{as\_of}, \C{as\_of\_valid}, \C{history}), counterfactual branching (\C{branch}), composable policy-gated effects (15-op catalog + handler stack + \C{policy\_eval}), boundary-checked contracts (\C{spec.conform}, \C{spec.parse}, \C{spec.explain\_for\_llm}), and content-addressed homoiconic plan ASTs with byte-identical round-trip (\C{plan.parse}, \C{plan.unparse}, \C{Node.id}). Two further capabilities are specified-but-not-shipped: multi-agent coordination (Txn), and live production steering (REPL); compositional skill learning sits on top of shipped Plan parse/validation with Phase-2.C execution and the 4-gate promotion still pending. The \C{:persistence.plan/node} and \C{:persistence.plan/skill} specs were already registered in the Phase 1 registry --- a deliberate parse-don't-validate move that froze the data shape ahead of the code that would consume it (\sectionref{sec:specs}); Phase 2.B confirms that discipline by shipping a parser that round-trips byte-identically to EDN and conforms against the pre-committed spec. We call this reporting discipline out because the deliberate gap between \emph{``substrate shipped''} and \emph{``vertical modules shipped''} is itself part of the methodological contribution: \textbf{the substrate is claimed once, and each derived capability is a property we can check against the log, not an engineered feature per agent.} A subsequent v0.4.0a1 substrate-primitives release (2026-04-26) hardened three substrate seams without changing any proposition: a typed \C{Provenance} schema with model-pluggable shape, a \C{Dispatcher} class layering handler-per-tag dispatch over the existing AST walker, and a \C{ProjectionAdapter.fork(branch\_id)} Protocol method that closes the projection-side counterpart to \C{branch()}. \C{PLAN\_CANONICAL\_VERSION} stays at 1; Proposition~\ref{prop:plan}'s content-addressing identity is unchanged.

\textbf{Neurosymbolic positioning.} Persistence is neurosymbolic not as a label but as a design principle: \emph{neural agency} (LLM decisions, embeddings, generative skills) operates over an explicitly \emph{symbolic substrate} (bitemporal datom queries with a Datalog-shaped surface, EDN AST grammars for plans, policy-as-data, Malli-style specs). The neural layer handles the open-ended; the symbolic layer handles accountability, composition, and guarantees. Concretely, Phase 1 ships four symbolic-substrate pieces and we deliberately scope two further symbolic capabilities \emph{out} of Phase 1 to keep the boundary between shipped artifact and aspiration sharp:

\begin{itemize}
\item \textbf{Bitemporal datom queries with a Datalog-shaped surface.} \C{as\_of}, \C{validAsOf}, \C{history}, \C{branch} are relational queries over the $(e, a, v, \tau, \dots)$ tuple store; Phase 1 implements them as Python comprehensions over the log (\C{src/persistence/fact/db.py}), with a Datalog rule engine adjacent future work.
\item \textbf{EDN AST grammars for plans.} \C{:persistence.plan/node} and \C{:persistence.plan/skill} are registered in the Phase 1 spec registry; the Phase 2.B Plan module ships parse / unparse / walk / spec-validation over a uniform \C{[tag \{attrs\} *children]} AST across all 16 node kinds with 128-bit content-addressing and (as of v0.3.0a1) a static-plus-manifest coercion registry that lets \C{datetime}, \C{bytes}, \C{Decimal}, \C{UUID}, \C{frozenset}, and \C{edn\_format.Symbol} participate in canonical hashing.
\item \textbf{Policy-as-data.} Shipped (\C{policy\_eval.py}, \textasciitilde{}200 LOC); guards, allow-lists, and rate limits are themselves datoms, queryable bitemporally and rewindable on a branch.
\item \textbf{Malli-style specs with LLM self-healing hints.} \C{persistence.spec} ships 186 tests and an \C{explain\_for\_llm} surface for parse-don't-validate boundary contracts (\sectionref{sec:specs}).
\end{itemize}

The two items we explicitly do \emph{not} claim as Phase 1 shipped are a Datalog rule engine and Z3-discharged \C{verify} leaves following Proof-of-Thought~\citep{fan2024proofofthought}; both are adjacent systems with compatible primitives, scheduled as future work (\S7.4 enumerates them in detail). This separation matters for a NeSy reviewer: the substrate's formal properties hold on what is shipped, and the future work is positioned as additive rather than load-bearing. The framing contrasts both with pure-neural approaches (prose CoT, embeddings-only memory) that surrender formal properties, and with pure-symbolic approaches that surrender expressiveness.

\textbf{Outline.} \sectionref{sec:related} surveys the five beachheads Persistence draws from. \sectionref{sec:thesis} states the six substrate invariants and maps them to Phase-1 shipped / Phase-2 designed boundaries. \sectionref{sec:formal} formalizes the substrate --- datoms (\sectionref{sec:datoms}), effect handlers and the well-formedness property (\sectionref{sec:effects}), the Merkle-hashed audit chain (\sectionref{sec:audit}), plans (\sectionref{sec:plans}), trajectories and replay with the byte-identical NO-OP invariant (\sectionref{sec:replay}), transactions (\sectionref{sec:transactions}), and specs with the self-healing LLM contract (\sectionref{sec:specs}). \sectionref{sec:impl} describes the reference implementation. \sectionref{sec:eval} presents a Reproduction Plan for the four evaluations --- camera-ready will replace [TBD] cells with measurements. \sectionref{sec:discussion} discusses limitations, privacy architecture, adoption path, and the neurosymbolic framing. \sectionref{sec:conclusion} concludes.

\section{Related Work}\label{sec:related}

\subsection{Agent memory}\label{sec:memory}

Agent memory has evolved rapidly. Vector-store RAG gave way to graph-structured memory (Graphiti, Neo4j-based agent graphs) as the benefits of relational traversal became clear. Zep~\citep{kurtic2025zep} introduced bitemporal edges with \C{valid\_at} and \C{invalid\_at} timestamps, with reported accuracy lifts over MemGPT on LongMemEval and substantial latency reductions; Memento~\citep{n1n2026memento} reached comparable figures with a similar design. (Exact percentage claims here will be cross-checked against the primary sources in the camera-ready.) Mem0 and A-Mem pursued a different direction --- \emph{mutable} memory that evolves via LLM-driven consolidation --- which~\citet{hannecke2026ssgm} showed produces unbounded drift without a formal write-gate. The Stability and Safety of Governed Memory framework proves $O(N \cdot \varepsilon_{\text{step}})$ bounded drift in systems with an append-only episodic ledger paired with a mutable projection, vs.\ unbounded drift in pure-mutable systems.

Persistence sits on the immutable end of this spectrum, with a critical extension: it treats \emph{every} datom --- not only memory facts, but audit entries, plan-AST nodes, skill versions, transaction commits --- as bitemporal. Memory is one consumer; the substrate serves all of them.

\subsection{Algebraic effects in LLM programming}\label{sec:effects-rw}

Algebraic effects, long studied in functional programming (Koka, Eff, Links), were recently brought to LLM scripting by~\citet{wang2025composable} and the Pangolin language~\citep{cheng2025pangolin}. Both separate \emph{what} an effect is from \emph{how} it is handled, enabling stacked interceptors for retry, caching, multi-shot sampling, and observability. Neither system, however, integrates effect handlers with a temporal memory substrate, nor with policy-as-data, nor with a formal audit chain. Persistence does: when the audit handler wraps the full catalog (the default configuration for regulated deployments), every effect emits a datom; every handler is policy-reviewable; every run is replayable. \sectionref{sec:audit} formalizes the Merkle-hashed audit contract.

\subsection{Structured agent programming}\label{sec:structured}

DSPy~\citep{khattab2023dspy} treats agents as declarative programs of typed modules with signatures, compiled by optimizers (BootstrapFewshot, MIPROv2) that search over instruction and demonstration slots. SmolAgents (Hugging Face, 2024) argues the opposite extreme --- code \emph{is} the best action language because LLMs already know Python. Open Agent Specification (2025) and declarative LLM-workflow DSLs formalize agent workflows as DAGs.

The symbolic-planning and knowledge-representation literature deserves a citation here. LLM+PDDL integrations~\citep{liu2023llmp,valmeekam2023planbench} argue for routing LLMs through classical planners; the same line --- articulated as direct evidence of unsoundness --- runs through~\citet{valmeekam2022plansurvey} and~\citet{kambhampati2024canllm}, who repeatedly argues that LLMs-as-planners are unsound without a symbolic verifier. Distinct from the LLM+PDDL line, the neurosymbolic reasoning literature --- Logic Tensor Networks~\citep{badreddine2022ltn}, DeepProbLog~\citep{manhaeve2018deepproblog}, and NeuPSL~\citep{pryor2023neupsl} --- integrates differentiable neural components with symbolic inference at a finer granularity than Persistence targets; our contribution is orthogonal: a substrate for agent-level symbolic scaffolding with LLM leaves, not a neural-symbolic reasoning hybrid at the inference layer. Persistence occupies neither pole: it is neither pure-LLM planning nor pure-PDDL verification, but a homoiconic AST in which leaf nodes may be LLM calls, symbolic code, or external planners --- committing to the AST shape via \C{:persistence.plan/node} without committing to the leaf implementation. The symbolic substrate carries the structural guarantees (content-addressing, descendant propagation, spec conformance); the leaves stay free to be neural, procedural, or SMT-discharged.

Persistence's Plan module is designed to reconcile these: the \emph{skeleton} of a plan is a symbolic EDN AST (editable, optimizable, compositional); the \emph{leaves} are prose or code (expressive, LLM-native). The Plan module itself is Phase 2; the \C{:persistence.plan/node} spec, however, is already registered in the Phase 1 spec registry (\sectionref{sec:specs}), committing to the AST shape before the evaluator ships.

\subsection{Skill libraries and continual learning}\label{sec:skills}

Voyager~\citep{wang2023voyager} demonstrated that LLM agents can accumulate executable skill libraries indexed by embeddings. Memento-Skills~\citep{memento2026skills} extended this with Read-Write Reflective Learning, framing memory updates as active policy iteration. Voyager promotes skills on their first successful use (\citealp{wang2023voyager}, \S3.3), without a statistical threshold; in noisy task distributions this produces a long tail of single-use skills.

Persistence's Plan module is designed to enforce a stricter four-gate promotion criterion: $\geq 3$ uses, $\geq 2$ distinct parent plans, $\geq 0.8$ rolling success rate, and retrieval round-trip consistency. Skills are to be versioned (\C{@v3}), content-addressed, and immutable; improvement produces a new version with a parent pointer, preserving lineage as in Git. (Plan module is Phase 2; the \C{:persistence.plan/skill} spec is registered in Phase 1 and encodes this promotion contract.)

\subsection{Counterfactual replay}\label{sec:counterfactual-rw}

CAMO~\citep{anon2026camo} introduced \emph{simulator-internal counterfactuals} with aligned randomness: paired rollouts with identical seeds, identical prefix state, and a single-variable intervention. AgentHER~\citep{anon2026agenther} adapted Hindsight Experience Replay~\citep{andrychowicz2017her} to LLM trajectories. AgenTracer~\citep{anon2025agentracer} uses counterfactual fault injection for multi-agent debugging. Abduct-Act-Predict~\citep{anon2025aap} operationalizes Pearl's causal ladder for LLM agents.

Persistence replay is stronger on the precise axis these papers care about. Because all state is immutable bitemporal datoms, ``aligned randomness'' is not a simulator hack but a substrate property. For the NO-OP intervention case, Phase 1's replay engine produces a trajectory whose canonical hash is \emph{byte-identical} to the factual trajectory's --- checked by \C{trajectory\_hash(cf) == trajectory\_hash(factual)} in \C{tests/replay/test\_determinism.py}. This is a stronger guarantee than the statistical-fidelity claims CAMO aspires to, because it holds at the level of the canonical serialization, not the level of outcome-delta distributions. The extension from toy agents to LLM trajectories (via per-step rng-consumption recording rather than fixed-draw-count \C{\_advance\_rngs\_to\_match}) is Phase 2 work; \sectionref{sec:replay} states Proposition~\ref{prop:noop} with this antecedent explicit.

\subsection{Positioning}\label{sec:positioning}

Table~\ref{tab:capability-matrix} places Persistence relative to the beachheads. Shipped Phase-1 capabilities are marked $\bullet$; Phase-2-designed capabilities are marked $\circ$.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{%
\scriptsize
\begin{tabular}{lccccccc}
\toprule
Capability & Zep/Graphiti & Pangolin & DSPy & Voyager & CAMO & \makecell{Persistence \\ (Phase 1 shipped)} & \makecell{Persistence \\ (Phase 2 designed)} \\
\midrule
Bitemporal memory                              & $\bullet$ &           &           &           &           & $\bullet$ &            \\
Effect handler stack                           &           & $\bullet$ &           &           &           & $\bullet$ &            \\
Merkle-hashed audit chain                      & partial   &           &           &           &           & $\bullet$ &            \\
Counterfactual replay (byte-identical NO-OP)   &           &           &           &           & partial   & $\bullet^*$ &          \\
Boundary specs + LLM self-healing hints        &           &           & partial   &           &           & $\bullet$ &            \\
Declarative plan AST                           &           &           & $\bullet$ &           &           &           & $\circ$    \\
Plan-AST optimization                          &           &           & partial   &           &           &           & $\circ$    \\
Skill library (4-gate promotion)               &           &           &           & $\bullet$ &           &           & $\circ$    \\
Multi-agent STM                                &           &           &           &           &           &           & $\circ$    \\
Live production REPL                           &           &           &           &           &           &           & $\circ$    \\
Regulator-replay fidelity                      &           &           &           &           &           & \multicolumn{2}{c}{[designed --- see \sectionref{sec:eval}]} \\
\bottomrule
\end{tabular}%
}
\caption{Capability coverage across related systems vs.\ Persistence (Phase 1 shipped, Phase 2 designed). $\bullet$ = supported; $\circ$ = designed-but-not-shipped; ``partial'' = covered with caveats. The asterisk on Persistence Phase-1 NO-OP replay denotes the toy-agent-vs-LLM-leaf footnote (see \sectionref{sec:replay}).}
\label{tab:capability-matrix}
\end{table}

$^*$Byte-identical NO-OP replay in Phase 1 is verified for toy agents with fixed rng-consumption; the per-step-rng-recording extension for LLM leaves is Phase 2 (\sectionref{sec:replay}, \S6.2).

\section{The Persistence Thesis}\label{sec:thesis}

We state six substrate invariants. Invariants 1, 2, and 5 are fully realized in Phase 1; invariant 3 is realized at the \emph{shape + parse + validate} layer by Phase 2.B (with the execution runtime in Phase 2.C); invariants 4 and 6 remain frozen at the spec-registry boundary with runtime implementation scheduled for Phase 2.

\begin{enumerate}
\item \textbf{Every fact is immutable, temporal, content-addressed.} Memory, audit, plan, skill, transaction --- all datoms. Provenance, counterfactual branching, and audit trails are derived, not engineered. \emph{[Phase 1: shipped --- \C{persistence.fact}.]}
\item \textbf{Every action is an effect.} LLM calls, tool calls, memory writes, and side effects pass through a composable handler stack. Policy, safety, caching, dry-run, and retry are handler instances. \emph{[Phase 1: shipped --- \C{persistence.effect}, 15-op catalog.]}
\item \textbf{Every plan is an EDN AST.} Agents read, edit, and evolve their own plans. Skills are named AST subtrees promoted under statistical evidence. Plan optimization is structural search. \emph{[Phase 1: \C{:persistence.plan/node} and \C{:persistence.plan/skill} specs registered (\C{src/persistence/spec/\_canonical.py}). Phase 2.B (shipped, v0.2.0a1): \C{persistence.plan} module with parse / unparse / walk / spec-validation over a uniform \C{[tag \{attrs\} *children]} AST across all 16 node kinds; 128-bit content-addressing via \C{sha256(canonical(node))[:32]}; byte-identical parse/unparse round-trip pinned by hypothesis property at \C{max\_examples=200}. Execution runtime (effect dispatch, optimizer ladder, 4-gate skill promotion) is Phase 2.C.]}
\item \textbf{Every shared state change is a transaction.} STM coordinates multi-agent belief updates without locks or message passing. \emph{[Phase 1: design frozen in \C{docs/agent*-spec.md}; Txn-module runtime is Phase 2.]}
\item \textbf{Every LLM boundary has a spec.} Parse-don't-validate at trust boundaries. Generative testing from specs. Self-healing via spec-error feedback. \emph{[Phase 1: shipped --- \C{persistence.spec}, 186 tests.]}
\item \textbf{Everything is REPL-live.} Inspection, editing, rewind, and speculative branching apply to running production agents, not only development loops. \emph{[Phase 1: rewind semantics subsumed by \C{as\_of} on the shipped Fact module; REPL-module runtime is Phase 2.]}
\end{enumerate}

\textbf{Thesis.} These six invariants, implemented together as a single runtime with a shared substrate, produce a system where accountability, replay, counterfactuals, composable safety, compositional skill promotion, coordinated multi-agency, and live production steering are \emph{properties of the substrate} --- not features engineered separately per agent. Phase 1 and the first Phase-2 increment together demonstrate five of these as shipped properties: accountability via the Merkle-hashed audit chain on the Fact log; replay via byte-identical NO-OP on the Replay engine; counterfactual branching via \C{branch()} on the Fact store; composable safety via the Effect handler stack with a machine-checkable well-formedness property; and homoiconic plan-as-data via the Phase 2.B Plan module's content-addressed parse/walk/spec-validation surface. Phase 2.C ships the remaining two capabilities (multi-agent coordination via Txn, live production steering via REPL) plus Plan-execution (optimizer ladder and 4-gate skill promotion) on top of the same substrate, with no substrate schema changes required --- that non-requirement is the testable form of the thesis.

\section{Formalization}\label{sec:formal}

The formalization below tracks the Phase 1 reference implementation. Where a claim depends on Phase 2 work, we mark it explicitly.

\subsection{Datoms and bitemporal queries}\label{sec:datoms}

A \textbf{datom} is an eight-tuple:

\begin{equation}\label{eq:datom}
d = \langle e,\ a,\ v,\ \tau,\ \tau_{sys},\ \nu_{from},\ \nu_{to},\ \omega \rangle
\end{equation}

where $e$ is an entity identifier, $a$ a namespaced attribute, $v$ an EDN value, $\tau$ a monotonic transaction id, $\tau_{sys}$ the ingestion (system) time, $\nu_{from}$ and $\nu_{to}$ the valid-time interval, and $\omega \in \{\text{assert}, \text{retract}\}$. A provenance record $\pi$ accompanies each datom; v0.4.0a1 ships its typed schema as the \C{Provenance} \C{TypedDict} (\C{total=False}) over seven known keys --- \C{source}, \C{tx\_time}, \C{handler\_id}, \C{canonical\_call}, \C{parent\_provenance\_hash}, \C{superseded\_by\_tx}, and a free-form \C{extra} map for caller-defined fields. The schema is deliberately model-pluggable: \C{handler\_id} names the registered effect handler that produced the datom, \C{canonical\_call} is a content hash of the call site (model + prompt + tools + params), and there is no \C{:provider} field --- the neural model is a plug-in, not embedded in the substrate's shape. An optional pointer \C{invalidated-by} links a superseded datom to the transaction that superseded it. Phase 1 seals provenance with a SHA-256 content hash. Cryptographic per-transaction signatures (ed25519) are Phase 2 work and are discussed as a privacy-posture extension in \S7.2.

The database $D$ is a finite set of datoms. We define the core queries:

\begin{itemize}
\item $\text{asOf}(D, t) = \{d \in D \mid \tau_{sys}(d) \leq t\}$
\item $\text{validAsOf}(D, t) = \{d \in D \mid \omega(d) = \text{assert} \land \nu_{from}(d) \leq t < \nu_{to}(d)\}$
\item $\text{history}(D, e) = \{d \in D \mid \text{entity}(d) = e\}$, sorted by $\tau$
\item $\text{branch}(D, t, \Delta) = \text{asOf}(D, t) \cup \Delta$, where $\Delta$ is a set of hypothetical datoms
\end{itemize}

Projection-layer counterfactuals are supported via \C{ProjectionAdapter.fork(branch\_id)} (v0.4.0a1; \C{DictProjection.fork()} reference impl) --- a Protocol method returning a fresh empty adapter that the caller populates by driving \C{rebuild()} against a branched DB. This closes a previously-implicit assumption that a counterfactual run could observe ordinary \C{apply()} writes from its parent's projection.

\begin{proposition}[Branch is a logical operation over the shipped store]\label{prop:branch}
\C{branch(D, t, $\Delta$)} returns a new \C{DB} value backed by a fresh in-memory store seeded with \C{asOf(D, t)} and extended with $\Delta$; writes to the branched value cannot leak back into the parent store. \emph{Complexity:} on the Phase 1 \C{InMemoryStore} reference implementation (\C{src/persistence/fact/db.py}), materialization is $O(|D|)$ in the seed snapshot plus $O(|\Delta|)$ in the hypothetical additions. \emph{Phase 2 upgrade:} under a persistent hash-array-mapped-trie (HAMT) backing store, the seed step reduces to $O(|\Delta| \log |D|)$ via structural path-copy; the \C{Store} Protocol boundary makes this a drop-in replacement requiring no change to \C{branch}'s interface.
\end{proposition}

\emph{Remark.} The Phase 1 complexity is honest rather than ideal: Python lists and dicts are not persistent data structures, and we chose to ship a correct list-backed implementation rather than pull in \C{pyrsistent} / \C{immutables.Map} for Phase 1. The \emph{isolation} property --- branched writes do not mutate the parent --- holds unconditionally on the shipped code and is the load-bearing property for counterfactual reasoning.

\subsection{Effects, handlers, and well-formedness}\label{sec:effects}

An effect signature is a typed pair $\langle \text{op}: \kappa,\ \text{args}: \sigma_{in} \rightharpoonup \sigma_{out} \rangle$. The minimum catalog $K$ is:

\begin{align*}
K = \{ & \C{llm/call},\ \C{tool/call},\ \C{mem/read},\ \C{mem/write},\ \C{decide},\ \C{ask-user}, \\
       & \C{emit-artifact},\ \C{sleep},\ \C{random},\ \C{env/read},\ \C{net/fetch},\ \C{secret/use}, \\
       & \C{cost/charge},\ \C{clock/now},\ \C{audit/emit} \}
\end{align*}

A \textbf{handler} $h$ is a partial function on operations with continuation semantics:

\begin{equation}\label{eq:handler}
h : (\kappa,\ \sigma_{in},\ k : \sigma_{out} \to \alpha,\ \mathrm{ctx}) \to \alpha
\end{equation}

A handler stack $H = [h_1, h_2, \dots, h_n]$ with $h_1$ outermost dispatches operation $\kappa$ to the outermost handler whose domain contains $\kappa$; that handler invokes the continuation $k$, which delegates to the remaining stack.

\begin{proposition}[Well-formedness; machine-checkable on the shipped runtime]\label{prop:wellformed}
A stack $H$ over catalog $K$ is well-formed iff for every $\kappa \in K$, at least one handler above the raw base handles $\kappa$. The shipped \C{Runtime.is\_well\_formed(catalog)} (\C{src/persistence/effect/runtime.py}) decides this property in $O(|H| \cdot |K|)$ time; \C{Runtime.uncovered\_ops(catalog)} returns the witness set. At runtime, \C{Runtime.perform(op, \dots)} raises \C{Unhandled} when no handler covers $\kappa$ --- the property is not merely asserted but \emph{enforced on every call}.
\end{proposition}

This is the paper's strongest formal contribution on the Phase-1 artifact: a decidable, linear-time, machine-checked completeness property over the runtime's neurosymbolic interface, with runtime enforcement via \C{Unhandled}. The check is exercised in the Phase 1 test suite. Properties above the effect layer (audit chain integrity \sectionref{sec:audit}, replay determinism \sectionref{sec:replay}) build on well-formedness of the deployed stack; policy composition is \emph{convenient} to express once well-formedness holds but is not itself a substrate invariant (\sectionref{sec:audit} treats audit-universality as a stack-configuration contract, not a substrate property).

Handlers compose \emph{masked} (following Koka's named/masked semantics): $\text{mask}_a(h)$ prevents $h$ from intercepting operations of attribute $a$ within its own body, eliminating re-entrant loops when, e.g., a policy handler itself needs to call the LLM.

\subsection{The Merkle-hashed audit chain}\label{sec:audit}

When the \C{audit} handler wraps a set of operations $W \subseteq K$, each effect on $\kappa \in W$ emits a datom recording (op, args, verdict, provenance, \C{prev\_hash}) before the continuation fires. The chain is built by taking \C{sha256(canonical\_serialize(entry) || prev\_hash)} at each link. Phase 1 ships:

\begin{itemize}
\item \C{make\_audit\_handler(wraps, \dots)} with configurable $W$ and a default of \C{("llm/call",)}.
\item \C{verify\_chain(entries) $\to$ bool}, which re-derives each \C{prev\_hash} from the canonical serialization and detects field-mutation tamper.
\item \C{audit\_entry\_to\_datom(entry) $\to$ datom-shaped record} that flows the audit entry into the Fact log.
\item \emph{(v0.4.0a1)} \C{audit\_entry\_to\_datom} additionally writes a \C{parent\_provenance\_hash} alias alongside the existing \C{:prev-hash} keyword --- both keys point to the same chain hash. This bridges audit-entry datoms into the typed \C{Provenance} schema (\sectionref{sec:datoms}) so that \C{DB.causal\_history(e)} walks audit chains and ordinary fact-log derivation chains under one query primitive. Tamper-evidence (Proposition~\ref{prop:tamper}) is preserved by construction: \C{verify\_chain} reads \C{:prev-hash} unchanged, and the new alias is an additional emitted field, not a replacement.
\end{itemize}

\textbf{Integrity contract.} \C{verify\_chain} detects any single-field mutation inside an entry (tested). Deletion/reorder coverage is flagged in the Round-1 rigor review and is a hardening target for Round 2. Authenticity --- proving \emph{who} signed --- is distinct from integrity and is not claimed for Phase 1: the current \C{signature} slot stores a SHA-256 content hash, and per-transaction ed25519 signing is Phase 2 work (\S7.2).

\begin{proposition}[Audit-chain immutability]\label{prop:tamper}
For any audit-chain $C = \langle e_0, e_1, \dots, e_n \rangle$ produced by \C{make\_audit\_handler}'s Merkle-hashed chain-append clause, \C{verify\_chain(C) = True} iff for all $i$, $e_i.\text{id} = \text{sha256}(\text{canonical}(e_i.\text{fields} \setminus \{\text{id}\}))$ and $e_i.\text{prev\_hash} = e_{i-1}.\text{id}$ --- i.e. no entry has been mutated, deleted, reordered, or truncated from the middle. This is exercised end-to-end by \C{tests/effect/test\_audit.py::test\_tampering\_an\_entry\_breaks\_the\_chain}, \C{test\_deleting\_an\_audit\_entry\_breaks\_the\_chain}, and \C{test\_reordering\_audit\_entries\_breaks\_the\_chain}; tail-truncation is allowed by construction (\C{test\_truncating\_audit\_entries\_from\_tail\_preserves\_chain}) and must be detected by regulators comparing a separately-recorded expected length.
\end{proposition}

\textbf{Universality contract.} ``Every effect emits a datom'' is an invariant of the deployed stack, not the substrate: it holds when and only when the audit handler's $W$ covers the full catalog. Phase 1 exposes \C{Runtime.is\_well\_formed(catalog)} to check coverage of $W$ against $K$; a \C{Runtime.assert\_universal\_audit} hardening is scheduled for Round 2. For regulated deployments (Case A in \S6.5), the configured stack wraps all 15 ops.

\subsection{Plans (Phase 2.B shipped: parse/walk/spec-validation; execution Phase 2.C)}\label{sec:plans}

A plan is a labeled tree where internal nodes are \emph{control operators} ($\text{seq}$, $\text{par}$, $\text{choice}$, $\text{loop}$, $\text{race}$, $\text{let}$, $\text{branch}$) and leaves are \emph{effect invocations} ($\text{tool-call}$, $\text{llm-call}$, $\text{code}$, $\text{checkpoint}$) or \emph{cognitive operators} ($\text{reflect}$, $\text{verify}$, $\text{call-skill}$). Every node $n$ carries a content identity $\text{id}(n) = \text{sha256}(\text{canonical}(n))[:32]$ used as its Merkle-DAG handle. The \C{:persistence.plan/node} spec is registered in \C{src/persistence/spec/\_canonical.py} (Phase 1); the Phase 2.B \C{persistence.plan} module (v0.2.0a1, merged 2026-04-24) ships parse, unparse, depth-first walk, and spec validation against this shape. The shipped spec encodes each node as an EDN vector \C{[:tag \{attrs\} \& children]} rather than a map --- a deliberate Lisp-style choice that makes plan ASTs \emph{homoiconic}: a plan literal is indistinguishable from the EDN data that describes it, and the homoiconicity contract's allowed self-edits (read / splice / compose / rewrite / fork / promote) reduce to list-splicing operations over well-typed vectors rather than reflective key-dispatch on a \C{:node/kind} field. This is what makes the parse-don't-validate methodology (\sectionref{sec:specs}) first-class rather than cosmetic.

\begin{proposition}[Plan content-addressing with descendant propagation]\label{prop:plan}
For any plan node $n = \langle \text{tag}, \text{attrs}, \text{children} \rangle$ constructed through the shipped \C{Node} constructor, the identity $\text{id}(n) = \text{sha256}(\text{canonical}(n))[:32]$ is a 128-bit prefix of a cryptographic hash over the canonical JSON serialization of $n$'s tag, its alphabetically-sorted attr map (with non-finite floats rejected at construction, non-string / colon-prefixed / empty / reserved (\C{id}, \C{:id}) attr keys rejected at construction, and user-supplied top-level \C{:id} additionally stripped at parse as defense-in-depth against external EDN hash-poisoning), and its children's ids recursively. Consequently, for any descendant modification of a tree $T$ at depth $d$ producing $T'$, every ancestor id on the root-to-leaf path differs between $T$ and $T'$. Property-checked at \C{max\_examples=200} on two headline properties (round-trip id preservation and descendant-mutation propagation) and \C{max\_examples=50} on four supporting properties by \C{tests/plan/test\_property.py}, with hypothesis strategies spanning all 16 node kinds, nested attr maps to depth 2, and float32 numeric values. \emph{Byte-identity of the canonical form under round-trip} is pinned by \C{test\_round\_trip\_preserves\_id}: for any hypothesis-generated $n$, $\text{id}(\text{parse}(\text{unparse}(n))) = \text{id}(n)$ --- a property that, before v0.2.0a3, was falsified by \C{Node}-constructed nodes carrying a user-supplied \C{id} attr that the parser silently stripped; the construct-vs-parse asymmetry is now closed by rejecting reserved keys at both sides. Spec conformance against \C{:persistence.plan/node} is enforced at parse time when \C{strict=True} and is exercised by 32 per-kind malformed parametrize cases (8 leaf-kinds $\times$ non-dict-attrs + 8 container-kinds $\times$ malformed-child + 16 kinds $\times$ wrong-tag-case); the 7 remaining \C{xfail} cases scope the Phase-2.C per-kind required-attr tightening and do \emph{not} mask the shipped claim. Birthday-collision probability at 1\% reaches $\approx 2.6 \times 10^{18}$ plans at the 128-bit width --- a width argued explicitly in the \C{CHANGELOG-plan.md} entry for the corresponding ship commit  % RESTORE: 'commit \C{190668a}' rather than left unstated.
\end{proposition}

v0.4.0a1 layers a handler-per-tag dispatch seam over the existing AST walker. The \C{Dispatcher} class (\C{persistence.plan.\_dispatch}) registers a handler per tag with raise-on-duplicate semantics by default and dispatches by walking the tree once --- walk order $\equiv$ dispatch order, property-tested with hypothesis at \C{max\_examples=50}. The walker module itself is renamed \C{\_walk.py} (was \C{\_interpret.py}; back-compat shim retained). This is the seam Phase 2.C will bind effect handlers into without requiring AST shape changes --- \C{PLAN\_CANONICAL\_VERSION} stays at 1.

\emph{Neurosymbolic reading.} Proposition~\ref{prop:plan} is the formal identity law for a \textbf{homoiconic neurosymbolic AST}. The skeleton (tag, attrs, children) is symbolic --- editable, optimizable, regulator-readable; the leaves (\C{:llm-call}, \C{:code}, \C{:reflect}) are neural. Descendant propagation means every structural edit of the symbolic layer produces a new ancestor-id path --- a verifiable identity surface that Phase-2.C plan optimizers (MIPROv2, MCTS-over-Plan) walk over the symbolic scaffold while leaves remain neural. To our knowledge, this is the first content-addressing identity law stated and property-checked for an LLM-leaf-bearing plan AST; the symmetric construct-and-parse rejection of reserved-id keys makes the identity a \emph{substrate} property rather than a parser contract.

The homoiconicity contract defines the agent's allowed self-edits: $\text{read}$, $\text{splice}$, $\text{compose}$, $\text{rewrite}$, $\text{fork}$, $\text{promote}$. Forbidden without explicit capability: editing $\text{verify}$ nodes (which would defeat any future proof-of-thought guarantees, per Proof-of-Thought~\citep{fan2024proofofthought}), modifying ancestors of the currently-executing frame (prevents self-suicide), introducing unbounded $\text{loop}$.

A \textbf{skill} is a quadruple $\langle \text{name},\ \text{version},\ \text{ast},\ \text{stats} \rangle$ with stats $\{\text{uses}, \text{success}, \text{cost}\}$, encoded by the \C{:persistence.plan/skill} spec. A subtree $T$ of plan $P$ is to be promoted to a named skill iff all four gates hold (Phase 2 runtime):

\begin{enumerate}
\item $\text{uses}(T) \geq 3$
\item $T$ appears in $\geq 2$ distinct parent plans
\item Rolling success rate of $T$ $\geq 0.8$
\item The LLM-generated docstring of $T$ retrieves $T$ as top-1 over the existing skill library.
\end{enumerate}

Gate (4) is the cheapest way to enforce semantic non-collision: if an LLM-written description of the skill cannot retrieve the skill itself from the library, the skill's function is not well-defined in the context of its peers.

\subsection{Trajectories and replay}\label{sec:replay}

A \textbf{trajectory} is an ordered sequence of effect datoms sharing a run-id, plus a seed vector $\sigma = \langle \sigma_{llm}, \sigma_{tool}, \sigma_{env} \rangle$. An \textbf{intervention set} is $I = [\langle \text{step},\ \text{field},\ \text{new-value} \rangle, \dots]$ --- a (possibly empty) list of per-step modifications to the counterfactual, sorted by \C{step}. The single-triple case is the Phase-1 default; the shipped replay engine and the \C{:trajectory/intervention} slot (registered as \C{seq\_of(:persistence.replay/intervention)}) both accept the multi-entry form.

The replay operator is:

\begin{align*}
\text{replay}(T, I) = T' \text{ where } & T'_i = T_i \quad \forall i < I.\text{step} \quad \text{(verbatim prefix)} \\
& T'_{I.\text{step}} = T_{I.\text{step}} \oplus I \quad \text{(intervention applied)} \\
& T'_i = \text{execute}(T_{i-1}, \sigma) \quad \forall i > I.\text{step} \quad \text{(re-executed suffix)}
\end{align*}

with effect calls in the suffix routed through a \emph{replay handler} that returns cached responses from the audit log, keyed by args-hash.

\begin{proposition}[Replay determinism --- stronger NO-OP form]\label{prop:noop}
If all non-determinism in the agent routes through effects in catalog $K$, \textbf{and} the per-step rng-consumption pattern is recorded, then $\text{replay}(T, I)$ produces a deterministic counterfactual trajectory that diverges from $T$ exactly at $I.\text{step}$. Phase 1 specializes and \emph{strengthens} this for the NO-OP case (an intervention that applies but does not change any value): \emph{(NO-OP byte-identity; tested.)} For a NO-OP intervention on a toy agent instrumented via the shipped replay engine, $\text{trajectory\_hash}(\text{replay}(T, I_{\text{noop}})) = \text{trajectory\_hash}(T)$ --- \emph{byte-identical} on the canonical serialization, not merely statistically close. Verified by \C{tests/replay/test\_determinism.py::test\_noop\_intervention\_produces\_byte\_identical\_trajectory}.
\end{proposition}

This strengthens CAMO's aspirational aligned-randomness claim in the direction NeSy reviewers will care about: the guarantee is a \emph{structural} property of the canonical-serialization hash, decidable byte-wise, not a statistical claim about outcome-delta distributions. Extending the property from a toy agent (whose per-step rng-consumption is fixed by construction in \C{conftest.py}) to real LLM trajectories requires replacing the current \C{\_advance\_rngs\_to\_match} heuristic with a recorded per-step rng-state vector. This is Phase 2 work and a prerequisite for the \S6.2 CAMO-style evaluation numbers in the camera-ready.

\subsection{Transactions}\label{sec:transactions}

A \textbf{ref} is a pair $\langle \iota, \tau_{\text{current}} \rangle$ where $\iota$ is an identity and $\tau_{\text{current}}$ the most recent transaction id. A transaction body is a thunk that reads and writes refs; commit emits an atomic set of datoms, CAS-ing on all read refs' $\tau$. On conflict, the transaction is retried. Composition is automatic: a transaction that calls another transaction inherits the outer's write set. The Txn-module runtime is Phase 2.

\subsection{Specs and the self-healing LLM contract}\label{sec:specs}

A \textbf{spec} $s$ is a predicate-generator pair $\langle p : \sigma \to \text{bool},\ g : \emptyset \to \sigma \rangle$. Parsing (conforming) a value $x$ against $s$ yields either a refined value or $\bot$ --- a true discriminated union in the shipped \C{\_conform} (\C{Conformed} \textbar{} \C{ConformError}, not silent coercion). The boundary contract is: business logic is typed against conformed values; raw inputs cannot reach it. Specs are first-class EDN data --- composable, diffable, versionable.

\textbf{Self-healing contract (shipped).} When conform fails, \C{spec.explain\_for\_llm(err)} returns a structured message containing the field path, the failure reason, and a Fix-clause-annotated hint. This is the agent-facing surface of the spec module: when an LLM produces a structurally-malformed tool call or plan node, the spec layer does not simply reject it but returns a prompt-ready correction hint that the agent can retry against. This formalizes a \textbf{conform $\to$ explain $\to$ retry} contract that turns spec violations into structured neurosymbolic self-correction, and is tested at \C{tests/spec/test\_llm\_errors.py}. We claim this as a concrete neurosymbolic contribution --- the symbolic layer emits hints that the neural layer can consume --- distinct from post-hoc error logging.

\textbf{Self-conforming producers (shipped).} Phase 1 specs are bidirectional contracts, not merely input guards. Every load-bearing producer --- \C{audit\_entry\_to\_datom}, \C{AuditEntry.to\_edn}, \C{Trajectory.to\_edn}, \C{datom\_to\_wire}, \C{wire\_to\_datom} --- calls \C{spec.conform(\dots)} against its own return value and raises \C{ValueError} on mismatch. A defect in a producer fails loudly at the producer's site, not later inside a consumer that read the bad wire form. This is a stronger invariant than consumer-side validation: the boundary contract is machine-checked at emission time, which means a paper-stated property about a wire shape is enforced by the code that emits it, not by downstream discipline.

\textbf{Forward-compatible spec-first commitment.} In Phase 1 we register \C{:persistence.plan/node} and \C{:persistence.plan/skill} in the spec registry \emph{before} the Plan module exists. This is a deliberate parse-don't-validate methodology choice: the data shape is locked before code depends on it, which lets Phase 1 and Phase 2 workers agree on the plan AST's structure without blocking on implementation order. The \C{:persistence.plan/node} spec is the commitment device; it enforces the paper's \sectionref{sec:plans} contract without requiring the runtime to exist.

\section{Implementation}\label{sec:impl}

Persistence is implemented across seven modules. The reference implementation is in Python for integration parity with existing agent stacks. Phase 1 ships four modules (Fact, Effect, Spec, Replay); Phase 2 ships the remaining three (Plan, Txn, REPL).

\subsection{Fact (Phase 1 --- shipped)}\label{sec:impl-fact}

The Fact module is designed for an append-only Postgres datom log with five covering indexes (EAVT, AEVT, AVET, VAET) plus a bitemporal VT-E range index and a log-ordered index (Datomic Index Model;~\citealp{tonsky2023datomic}). The shipped Phase 1 reference implementation ships a \C{Store} Protocol with two backends: \C{InMemoryStore} (for tests and the CLI demo) and \C{SQLiteStore} (for zero-ops persistent deployments). The Phase 1 SQL migration (\C{migrations/0001\_datom\_log.sql}) ships against SQLite 3.37+ today; the bitemporal datom wire form is migration-compatible, with a Postgres adapter planned for Phase 2. Zstd-compressed segments and content-addressed storage are explicitly Phase 2 work (per \C{CHANGELOG.md}).

Writes pass through a transactor that computes auto-retractions for cardinality-one attribute overwrites, preserving historical values with \C{invalidated-by} pointers. The projection surface is a \C{ProjectionAdapter} Protocol (\C{reset}, \C{apply}) with a reference in-process \C{DictProjection} in Phase 1; production Kuzu and mem0 projection adapters are Phase 2 work. A legacy-write \C{mem0\_adapter} interceptor is shipped in Phase 1 --- it emits a datom before delegating to an operator-supplied mem0 client --- but it is an \emph{interceptor} rather than a projection. In v0.4.0a1, \C{ProjectionAdapter} gains a \C{fork(branch\_id) -> ProjectionAdapter} Protocol method (with \C{DictProjection.fork()} as the reference impl): the fork returns a fresh empty adapter, and the caller drives \C{rebuild()} to populate it from the branched DB. This makes counterfactual reasoning a substrate-level property over both the fact log AND the projection --- the projection sink no longer leaks parent-DB writes into a counterfactual run. \C{DB.causal\_history(e, max\_depth=16)} ships in the same release as the substrate-level lineage query primitive (single-level extraction in v0.4.0a1; multi-level walking deferred to v0.5).

\textbf{Concurrent-writer safety.} \C{SQLiteStore.allocate\_and\_append(datoms)} runs the \C{MAX(tx) + 1} allocation and the row INSERTs inside a single \C{BEGIN IMMEDIATE} transaction; \C{InMemoryStore.allocate\_and\_append} does the same under its \C{threading.Lock}. This gives the \C{Store} Protocol a single atomic allocate-and-append primitive that the transactor routes through, closing the TOCTOU window that a prior \C{next\_tx()}-then-\C{append(\dots)} split would have left open under multi-writer load. The guarantee is exercised by \C{tests/fact/test\_concurrent\_transact.py}: 16 threads $\times$ 50 transacts under a \C{threading.Barrier} produce 800 unique tx ids with zero collisions. This is the concurrency invariant that Proposition~\ref{prop:noop} (\sectionref{sec:replay}, NO-OP byte-identity) implicitly depends on: without a race-free allocator, two concurrent replays sharing a backing store would produce colliding tx ids and break the trajectory-hash identity.

\textbf{Latency targets (Reproduction Plan).} Phase 1 tests check correctness, not latency. The paper's per-operation p95 targets --- \C{as\_of} $\leq$ 50 ms, \C{branch} $\leq$ 200 ms, \C{history(e)} $\leq$ 100 ms for entities with $\leq$ 1000 datoms --- are Phase-2 measurements over a persistent-trie backing store at 1M-datom scale. Phase 1 reference-implementation numbers over the \C{InMemoryStore} and \C{SQLiteStore} backends at 1k / 10k / 100k datom corpora will be reported in \sectionref{sec:eval} of the camera-ready; they are [TBD] in this draft.

\subsection{Effect (Phase 1 --- shipped)}\label{sec:impl-effect}

The Effect module implements a handler-stack runtime. Handlers are declared in EDN with clauses keyed by operation. The canonical stack for regulated domains is:

\begin{verbatim}
audit -> policy -> dry-run -> cache -> retry -> rate-limit -> raw
\end{verbatim}

Each handler is a pure function over the operation, arguments, and continuation; \C{audit} emits a datom into the Fact module, chaining via \C{prev-hash} for Merkle integrity. Policies are declarative EDN interpreted by a \textasciitilde{}200-line evaluator (\C{policy\_eval.py}, measured) supporting principal attributes, op-matching, and conditional effects (\C{:op=}, \C{:op-in}, \C{:mode=}, \C{:and}, \C{:or}, \C{:not}, \C{:contains?}, \C{:matches?}). The \C{Runtime.is\_well\_formed} and \C{Runtime.uncovered\_ops} functions (\sectionref{sec:effects}) are part of the public API.

\subsection{Spec (Phase 1 --- shipped)}\label{sec:impl-spec}

The Spec module provides Malli-equivalent functionality in Python: predicate-generator pairs, composable registry, conformance. Specs are EDN data. A generative testing harness produces example instances for every registered spec; failed conforms generate spec-error messages used as self-healing retry hints to the LLM (\C{spec.explain\_for\_llm}, \sectionref{sec:specs}). Specs attach to every boundary. Ten canonical specs are registered, including \C{:persistence.fact/datom}, \C{:persistence.effect/audit-entry}, \C{:persistence.replay/trajectory}, \C{:persistence.plan/node}, and \C{:persistence.plan/skill} --- the last two registered ahead of the Plan module.

\subsection{Replay (Phase 1 --- shipped)}\label{sec:impl-replay}

The Replay module records trajectories as sequences of effect datoms and implements the replay operator from \sectionref{sec:replay}. A replay handler intercepts effects during branching, returning cached responses indexed by args-hash. DPO pairs are to be extracted when paired trajectories differ in outcome beyond a configurable threshold: prefix must match exactly; suffix divergence feeds the \C{chosen}/\C{rejected} dataset automatically. \C{gen\_regression\_test(trajectory)} emits a pytest-source snapshot-test string that asserts against the loaded trajectory (this is honest snapshot-replay rather than agent-re-run; full regression re-execution is Phase 2). The byte-identical NO-OP determinism invariant (\sectionref{sec:replay} corollary) is the headline test result for this module.

\subsection{Plan (Phase 2.B --- shipped; Phase 2.C --- designed)}\label{sec:impl-plan}

The Plan module v0.2.0a1 ships a minimal parse/walk/spec surface over the uniform \C{[tag \{attrs\} *children]} EDN AST covering all 16 registered node kinds (\C{:seq}, \C{:par}, \C{:choice}, \C{:case}, \C{:loop}, \C{:race}, \C{:let}, \C{:branch}, \C{:ref}, \C{:tool-call}, \C{:llm-call}, \C{:code}, \C{:checkpoint}, \C{:reflect}, \C{:verify}, \C{:call-skill}). Four Python files (\C{\_ast.py}, \C{\_errors.py}, \C{\_parse.py}, \C{\_walk.py}) implement: a frozen-slots \C{Node} dataclass with content-addressed \C{id} (Proposition~\ref{prop:plan}), an \C{edn\_format}-backed parser with optional alias lowering (\C{:phase}/\C{:workstream} $\to$ \C{:seq}), a symmetric canonical unparser that round-trips byte-identically for canonical inputs, a depth-first \C{walk(node, visitor) -> list[str]} trace primitive, and \C{PlanSpecError}-wrapped conformance against \C{:persistence.plan/node}. Attr keys are plain strings internally (no leading colon), re-prefixed at the spec/emit boundary; user-supplied top-level \C{:id} is stripped at parse so that content-addressing is a function of the structural content alone. NaN/Inf in attrs are rejected at \C{Node.id} computation (\C{allow\_nan=False}) so that two semantically-distinct nodes cannot hash-collide via non-finite floats.

The v0.2.0a1 release passed an ARIS gate (R1 Correctness 8.2 / R2 Rigor 7.4$\to$8.8 after an 8-commit fix-pass / R3 Composability 8.3; R4 Research skipped for lack of external-system wiring at this scope), with the R2 fix-pass closing seven rigor MAJORs including a co-located \C{:id}-clobber defect and a 64$\to$128-bit identity-width argument. Subsequent v0.2.0a2 and v0.2.0a3 patches landed the R3 hardening items and closed a reserved-attr-key Proposition~\ref{prop:plan} round-trip falsifier discovered by hypothesis at \C{max\_examples=200}. 172 plan-module tests (7 xfailed) ship alongside the full-repo suite of \textbf{832 passed + 7 xfailed} at v0.4.0a1.

\textbf{Phase 2.C (designed, not shipped).} The execution runtime adds a \C{fold(node, reducer, init) -> (acc, trace)} primitive that supersedes \C{walk} for speculation, rollback, and checkpointing; an effect-dispatched evaluator emitting OpenTelemetry spans keyed by \C{node.id}; the optimizer ladder (MIPROv2 on \C{:llm-call} leaves delegating to DSPy, UCT-MCTS on \C{:choice} / \C{:branch} with Pareto dominance on \C{\{success, latency, cost\}}, evolutionary search on topology with LLM-synthesized mutations); and the 4-gate skill-promotion pipeline of \sectionref{sec:plans}. Finetuning is the final tier for stable hot skills with $\geq 1000$ uses.

\textbf{Hardening-track status honesty.} Four items shipped across v0.2.0a2 and v0.3.0a1 and are no longer future work: (i) a shared \C{persistence.spec.SpecError} base class so \C{PlanSpecError} composes with future sibling errors; (ii) a \C{:original-tag} escape hatch preserving aliased tags for MCTS/optimizer consumers; (iii) an explicit schema-evolution contract documenting how ids change when new default-valued attrs are introduced; (iv) a \C{Node.id} coercion registry for non-JSON-native types covering \C{datetime} / \C{date} / \C{bytes} / \C{Decimal} / \C{UUID} / \C{frozenset} / \C{edn\_format.Symbol}, shipped as a static-plus-manifest registry with \C{PLAN\_CANONICAL\_VERSION = 1} introduced --- coercion is id-time-only so \C{node.attrs} keeps the author's input while the canonical form sees the deterministic projection. Two items remain open and are Phase-2.C work: (i) the \C{fold()} primitive itself; (ii) substrate-wide 128-bit id-width consistency across Fact, Replay, and Plan. v0.4.0a1: \C{\_interpret.py} renamed to \C{\_walk.py} (back-compat shim retained); \C{Dispatcher} class added in \C{persistence.plan.\_dispatch} for handler-per-tag dispatch (see \sectionref{sec:plans}).

\subsection{Txn (Phase 2 --- designed)}\label{sec:impl-txn}

The Txn module will implement software transactional memory over refs backed by the Fact store. Commit semantics are designed to follow the Multiverse~\citep{bieniusa2026multiverse} hybrid of versioned and unversioned execution, enabling optimistic concurrency with automatic retry. The integration with Fact means every transaction is itself a datom; the entire multi-agent coordination history is queryable.

\subsection{REPL (Phase 2 --- designed)}\label{sec:impl-repl}

The REPL module will expose a capability-gated WebSocket API for live inspection, editing, rewind, and branching of running agents. The UI is an extension of an in-house thread-routing Claude Code client. Rewind resumes an agent from an earlier $\tau$ using the already-shipped \C{as-of} view; branch runs a speculative path in a sandboxed handler stack with dry-run forced. Production invariant: a REPL session cannot write to the authoritative fact store without a signed operator approval.

\subsection{System diagram}\label{sec:diagram}

\begin{figure}[ht]
\centering
\begin{verbatim}
  ┌─────────────────────────────────────────────────────────┐
  │  AGENT (business logic, domain-specific)                │
  └───────────────────────┬─────────────────────────────────┘
                          │ (perform :op args)
                          ▼
  ┌─── Effect ─── Spec ──┤ Txn [Phase 2] ├─────────────────┐
  │   handler stack, boundary contracts, STM commits        │
  └───────────────────────┬─────────────────────────────────┘
                          │ (emit datoms)
                          ▼
  ┌─── Fact ────────────────────────────────────────────────┐
  │   InMemory / SQLite log  ·  DictProjection              │
  │   [Phase 2: Postgres + Kuzu + mem0]                     │
  └─── ▲ ────────────────────────────────── ▲ ──────────────┘
       │ as-of / branch / history           │
       │                                    │
  ┌─── Replay ──────┐             ┌─── Plan [Phase 2] ──────┐
  │  trajectories   │             │  EDN AST + skills       │
  │  counterfactual │             │  MIPROv2 / MCTS / evo   │
  │  NO-OP identity │ ──evidence→ │  4-gate promotion       │
  └─────────────────┘             └─────────────────────────┘
                          ▲
                          │
  ┌─── REPL [Phase 2] ─────┴────────────────────────────────┐
  │   inspect / edit / rewind / branch (capability-gated)   │
  └─────────────────────────────────────────────────────────┘
\end{verbatim}
\caption{Persistence runtime stack. Phase 1 ships Fact / Effect / Spec / Replay; Phase 2 ships Plan execution / Txn / REPL.}
\label{fig:system-diagram}
\end{figure}

\section{Evaluation --- Reproduction Plan}\label{sec:eval}

The abstract submission (2026-06-09) reports the Phase-1 shipped artifact with formal properties (Proposition~\ref{prop:wellformed}; Proposition~\ref{prop:noop}) checked in the bundled test suite (832 passed + 7 xfailed, \C{pytest -q} from a clean clone). The numeric evaluations below are a \textbf{Reproduction Plan}: each subsection below names the harness, the dataset, the license, and the intended submission point (abstract vs.\ camera-ready). Tables marked [TBD] carry target numbers only; measured numbers land in the camera-ready (2026-07-20) once Phase 2 verticals and the per-step-rng-recording replay extension are in place.

\subsection{LongMemEval (camera-ready)}\label{sec:eval-longmem}

LongMemEval measures long-horizon memory fidelity across agent conversations. We will compare Persistence's Fact module against published results on the LongMemEval v1 corpus. The integration is straightforward in principle because our \C{as\_of\_valid} query is the same temporal filter Zep's bitemporal edges implement; the mem0 projection adapter (Phase 2) is the gating dependency for retrieval at scale, without which the harness falls back to linear log-scan (unusable beyond 10k conversation turns).

For the abstract, this subsection is qualitative; the camera-ready will ship numeric rows.

\begin{table}[ht]
\centering
\small
\begin{tabular}{lcc}
\toprule
System & Accuracy & p95 latency \\
\midrule
MemGPT          & [per primary source] & [per primary source] \\
Mem0            & [per primary source] & [per primary source] \\
Mem0g           & [per primary source] & [per primary source] \\
Zep / Graphiti  & [per primary source] & [per primary source] \\
Memento         & [per primary source] & [per primary source] \\
\textbf{Persistence (this work)} & \textbf{[TBD --- camera-ready]} & \textbf{[TBD --- camera-ready]} \\
\bottomrule
\end{tabular}
\caption{LongMemEval target table (camera-ready).}
\label{tab:longmemeval}
\end{table}

\textbf{Hypothesis (to test).} Persistence matches Zep's accuracy (we share the bitemporal model) with competitive latency (materialized projection is architecturally similar). The contribution is not a raw memory win but the substrate's composability with audit, replay, and plan storage.

\subsection{CAMO-style counterfactual fidelity}\label{sec:eval-camo}

This evaluation has two components. The \emph{byte-identical NO-OP} property (Proposition~\ref{prop:noop}) is \textbf{already testable on the Phase 1 artifact} --- \C{tests/replay/test\_determinism.py::test\_noop\_intervention\_produces\_byte\_identical\_trajectory} passes; we cite it directly in the abstract as the structural baseline stronger than CAMO's statistical aligned-randomness. The \emph{distributional CAMO protocol} --- 1000 paired rollouts with single-variable interventions, measuring prefix alignment (Hamming distance from factual), intervention faithfulness (probability the counterfactual at step $k$ reflects the intervention), and suffix variance over 100 re-replays --- requires (i) the per-step rng-state recording extension to the replay engine (Phase 2) and (ii) a meaningful LLM budget for stochastic agents. Both are feasible within the 2026-06-09 $\to$ 2026-07-20 camera-ready window if scoped to a single base agent.

\textbf{Reproduction Plan.} Abstract ships the NO-OP corollary. Camera-ready ships the full 1000-trajectory table on a toy agent with honest per-step rng recording and, if budget permits, a scaled-down 100-trajectory table on Claude Haiku as the stochastic agent.

\subsection{Regulator-replay benchmark --- 50 synthetic trajectories (novel)}\label{sec:eval-regulator}

We propose \textbf{regulator-replay fidelity}, a new benchmark. Given an agent decision trajectory produced in a regulated domain, can a third-party auditor --- given only the datom log and the plan AST --- deterministically reconstruct the production decision? The metric is the fraction of decisions whose reconstruction matches production byte-for-byte under \C{effect.canonical.canonical\_dumps} serialization.

\textbf{Scope (rescoped from the v0.1 draft).} For this paper we target a \textbf{50-trajectory synthetic project-finance corpus}, not a 200-trajectory production corpus. The synthesis pipeline runs a project-finance scoring agent (WACC, gearing, concession-fee, sector classification) over synthetic inputs, persists the resulting trajectories through the Phase 1 Fact module, and hands the log to an independent reconstruction script. The dataset generator, the reconstruction harness, and the 50-trajectory corpus will be released together under a permissive open-data license before camera-ready.  % RESTORE: 'CC-BY-4.0' Full production-scale evaluation is deferred to an extended-version companion paper.

\textbf{Why this rescope.} This is the paper's flagship novelty. A protocol proposal without numbers would collapse the contribution from ``we introduce and evaluate a new benchmark'' to ``we propose a benchmark.'' Fifty synthetic trajectories is the minimum tractable scope that keeps ``evaluate'' in the contribution.

\textbf{Symbolic-neural coupling.} Unlike pure-trace replay approaches~\citep{anon2026camo,anon2026agenther}, regulator-replay requires the \emph{symbolic plan AST at decision time} to be reconstructible alongside the neural effect trace; the benchmark measures this symbolic-neural joint reconstruction as a single fidelity metric. A reconstruction counts as faithful only when both the AST Merkle-id path and the effect-trace canonical form recover byte-identically from the log.

\begin{table}[ht]
\centering
\small
\begin{tabular}{lccc}
\toprule
Regulator-replay configuration & Factual deterministic & CF byte-identity @ NO-OP & Reconstruction fidelity \\
\midrule
50 synthetic PF trajectories, DictProjection & \textbf{[TBD]} & \textbf{[TBD --- target $\geq$ 99\%; H1]} & \textbf{[TBD]} \\
\bottomrule
\end{tabular}
\caption{Regulator-replay fidelity target (camera-ready). H1: see hypothesis below.}
\label{tab:regulator-replay}
\end{table}

\textbf{Hypothesis H1.} Reconstruction fidelity $\geq$ 99\% on the 50-trajectory synthetic corpus; deviations diagnostic of catalog well-formedness (non-determinism admitted into $K$ during trajectory synthesis) rather than substrate limits. The 1\% gap, where present, is attributable to a gap in the Phase-1 well-formedness check rather than the content-addressed audit chain.

\subsection{Plan-optimization comparison \emph{(removed from this paper)}}\label{sec:eval-removed}

The Plan-optimization comparison (HotpotQA, replayed-trade trajectories, project-finance scoring) requires the Plan module execution runtime, which is Phase 2.C. It is therefore out of scope for this paper and will be the core contribution of a Phase-2 companion paper once the Plan-execution module ships.

\subsection{Case studies --- one production deployment, three anonymized vignettes}\label{sec:eval-cases}

\textbf{Case B --- a production algorithmic-trading deployment (reported with numbers).}  % RESTORE: 'Adaptive Trader v2 (named deployment ...)' Case B exercises Persistence on a neurosymbolic trading agent where the \emph{playbook AST} (symbolic: entry conditions, regime classifiers, risk gates) composes with a \emph{neural decision layer} (Claude-as-sole-decision-maker) through an effect-captured non-determinism seam. Every decision writes a datom; every playbook version is content-addressed; every counterfactual replay walks the symbolic tree while re-sampling the neural leaves.

% RESTORE: verbatim block — case-name 'Adaptive Trader v2'; symbols 'BTCUSDT', 'ETHUSDT'
\begin{verbatim}
;; Case-B playbook AST (excerpt)
[:seq {:regime "trend-following"}
  [:llm-call {:tool "market-classifier" :model "claude-opus"}
    [:context {:window "4h" :symbols ["ASSET-1" "ASSET-2"]}]]
  [:verify {:gate "risk-limit" :max-pos-pct 0.20}]
  [:branch {:on :regime-classification}
    [:case {:when "trend"}
      [:llm-call {:tool "entry-sizer" :playbook "momentum-v7"}]]
    [:case {:when "range"}
      [:llm-call {:tool "entry-sizer" :playbook "mean-revert-v3"}]]]
  [:effect {:audit true :kill-switch :position-limit}]]
\end{verbatim}

Every node in this excerpt is content-addressed by Proposition~\ref{prop:plan}; the neural leaves (\C{:llm-call} nodes) compose with the symbolic control flow (\C{:seq}, \C{:branch}, \C{:case}, \C{:verify}) under the effect handler stack of \sectionref{sec:effects}. A single decision writes a datom whose payload includes the playbook's root \C{:id}, making the audit trail lineage-complete under structural edits.

Concretely, Case~B is a production algorithmic-trading deployment with a small live-capital cap and ARIS-reviewed risk governance. The factual dry-run baseline, recorded across 13 continuous days immediately preceding the Persistence migration, closed at profit factor 0.43 over 8 entries with negative PnL --- a clear NO-GO signal that motivated the redesign.  % RESTORE: 'Adaptive Trader v2 ... Binance Futures ... $400 USDT ... -$26.87 PnL' We instrument every decision through Persistence's effect handler stack (audit, kill-switch, position-limit, ARIS-gate, dry-run), persisting trajectories to the Fact module. A nightly cron runs the Replay module over each losing trajectory with an entry-delay wait-sweep $\{0, 5, 10, 15, 30, 60\}$ minutes. DPO-pair extraction (Phase 1 replay engine) is staged behind the per-step-rng-recording extension (Phase 2), and preliminary prompt-tuning results on the 8-trade baseline are therefore [TBD --- camera-ready]; the baseline numbers themselves are real. Case~B serves as the reference deployment throughout this paper because every Persistence invariant --- immutable decision trail, effect-captured non-determinism, EDN playbook AST, counterfactual-validated skill promotion --- is stress-tested against real-money risk. A forthcoming companion paper will report the extended live-capital evaluation.  % RESTORE: 'Adaptive Trader v2'

\begin{table}[ht]
\centering
\small
\begin{tabular}{lcc}
\toprule
Case~B & Pre-Persistence baseline & With Persistence (Phase 1) \\  % RESTORE: 'Adaptive Trader v2 (Case B)'
\midrule
Days of dry-run & 13 & [TBD --- camera-ready] \\
Entries & 8 & [TBD] \\
Profit factor & 0.43 & [TBD] \\
PnL & negative (small live-capital cap) & [TBD] \\
Audit-chain entries per decision & n/a (prose logs) & [TBD] \\
Regulator-replay fidelity on 8 baseline trajectories & n/a & [TBD --- target $\geq$ 99\%] \\
\bottomrule
\end{tabular}
\caption{Case~B baseline vs.\ Phase-1 Persistence-instrumented run.}  % RESTORE: 'Adaptive Trader v2'
\label{tab:case-b}
\end{table}

\textbf{Case A --- Project-finance assessment SaaS (anonymized vignette).} A bitemporal migration of a production assessment platform serving development finance institutions. The platform produces regulator-facing bankability scores for infrastructure projects; every assumption in the scoring pipeline (WACC, gearing, concession-fee, sector classification) must be reconstructible under audit. Phase 1 lands the Fact module as the authoritative scoring-assumption log; Phase 2 integrates Plan for the scoring pipeline AST. Numeric reporting deferred to a co-authored extended-version paper once the client IP-sharing agreement closes.

\textbf{Case C --- Insurance aggregator (anonymized vignette).} A licensed insurance comparator operating under a regulatory regime requiring full quote-issuance traceability. Bitemporal client state resolves compliance queries of the form \emph{``when did the agent learn of carrier price change X, and what quotes did it issue between the change and our ingestion?''} --- a bitemporal \emph{exclusive-OR} of valid-time and transaction-time, structurally impossible without both clocks. Phase 1 shipped; Plan-driven quote pipeline is Phase 2. Numeric reporting deferred pending licensure co-disclosure.

\textbf{Case D --- Hospitality operations (anonymized vignette).} A multi-tenant hotel operations agent connected via WhatsApp. Regression trajectories are auto-generated from customer feedback sessions on the production messaging channel; every pull request is designed to replay all known-failing trajectories through the new handler stack in CI. Phase 1 ships the trajectory capture; full regression re-execution is Phase 2 (\sectionref{sec:impl-replay}). Numeric reporting deferred pending operator-side deployment gate.

Case~B is the only deployment reported with numbers in this paper.  % RESTORE: 'Adaptive Trader v2' Case A, C, and D identities are withheld pending client-side co-authorship decisions; their reproducibility artifacts are available on request to NeSy program chairs under NDA.

\subsection{Reproduction posture}\label{sec:eval-posture}

\begin{itemize}
\item \textbf{Artifact.} The Phase-1 + Phase-2.B + v0.4.0a1 artifact is bundled with the paper submission. \C{pytest -q} from a clean clone runs the \textbf{832-passed / 7-xfailed} test suite in under one minute.
\item \textbf{Benchmark timeline.} \sectionref{sec:eval-longmem} (LongMemEval) and \sectionref{sec:eval-camo} (CAMO 1000-trajectory table) ship in the camera-ready (2026-07-20). \sectionref{sec:eval-regulator} (50-trajectory synthetic regulator-replay) ships in the camera-ready with the generator script. \sectionref{sec:eval-cases} Case B ships post-Persistence-migration numbers in the camera-ready.
\item \textbf{Licensing.} Licensing details deferred to camera-ready.  % RESTORE: 'Runtime: AGPL-3. Paper + benchmark harness + regulator-replay dataset: CC-BY-4.0.'
\item \textbf{Abstract submission scope.} At the 2026-06-09 abstract deadline, this section reports the formal properties (Proposition~\ref{prop:wellformed} and Proposition~\ref{prop:noop}) as \emph{already-checked on the shipped artifact}; all numeric tables carry [TBD] honestly.
\end{itemize}

\section{Discussion}\label{sec:discussion}

\subsection{Limitations}\label{sec:limitations}

Persistence Phase 1 imposes costs we disclose frankly:

\begin{itemize}
\item \textbf{Write latency} rises by a constant overhead per transaction due to transactor validation and auto-retraction. The exact overhead on the Phase 1 \C{InMemoryStore} and \C{SQLiteStore} backends is [TBD] (to be measured for the camera-ready latency table). Aspirationally, this is acceptable for agent workloads; unacceptable for voice-turn-latency inner loops, which we route to a fast ring buffer drained asynchronously. Per-transaction ed25519 signing --- which would add a further cryptographic-signing cost when enabled --- is Phase 2 work; no overhead figure is claimed for it until the signing path ships and is measured.
\item \textbf{Storage} on an append-only log is approximately $4\times$ a mutable-memory baseline, partially offset by content-addressing and (Phase 2) Zstd segment compression. Cold-tier eviction after 90 days (retain datoms, drop embeddings) mitigates further; Phase 1 ships the retention primitives, not the eviction scheduler.
\item \textbf{\C{branch} complexity} on Phase 1 is $O(|D|)$-copy (Proposition~\ref{prop:branch}), not $O(|\Delta| \log |D|)$. A persistent-trie backing Store is the Phase 2 upgrade path.
\item \textbf{Replay determinism} on Phase 1 is proven byte-wise only on a toy agent (Proposition~\ref{prop:noop}). The generalization to LLM trajectories requires per-step rng-state recording in the replay engine --- a targeted Phase 2 change that unblocks the \sectionref{sec:eval-camo} camera-ready numbers.
\item \textbf{Audit universality} is a configuration of the deployed stack, not an invariant of the substrate (\sectionref{sec:audit}). \C{Runtime.assert\_universal\_audit} is on the Round-2 hardening list.
\item \textbf{Policy-as-EDN} presumes Clojure/EDN literacy of the operator. The evaluator itself is language-agnostic at the data level, and JSON is a subset of EDN, so interoperability is straightforward.
\item \textbf{Substrate lock-in.} Adopting Persistence means adopting a specific opinionated model of agent state. The migration path is additive (\sectionref{sec:adoption}), but eventual consolidation around datoms is the intended end state.
\end{itemize}

\subsection{Privacy architecture}\label{sec:privacy}

Persistence is designed for local-first deployment. The authoritative datom log runs on operator-controlled infrastructure; all projections are regenerable. Provenance is \textbf{content-hashed per-datom (SHA-256) in Phase 1; per-transaction ed25519 signing is Phase 2 work} once key management (\C{sign\_transaction(tx\_id, tx\_hash, key) -> bytes}, verified inside \C{verify\_chain}) ships. A \C{:privacy :local} attribute on \C{:llm-call} nodes routes inference to local models (Qwen, DeepSeek, Llama), bypassing cloud vendors. Telemetry is emitted to an operator-controlled OpenTelemetry collector, with no SaaS egress. Skill-library entries carry a \C{:visibility :private} attribute enforcing never-to-cloud semantics.

For regulated deployments, the system is fully auditable with zero third-party data exposure once the audit handler wraps the full catalog (\sectionref{sec:audit}). A plausible commercial deployment puts the runtime inside a client's VPC, stores the log in client-controlled Postgres, and exposes only the REPL and metrics to the vendor. Cryptographic \emph{authenticity} of the audit chain (who signed the transaction) ships with ed25519 in Phase 2; Phase 1 provides \emph{integrity} (the chain detects tampering given itself) via \C{verify\_chain}.

\subsection{Adoption path}\label{sec:adoption}

Persistence is explicitly additive. An existing agent stack migrates in three phases:

\begin{enumerate}
\item \textbf{Datom log (2 days, non-breaking).} Stand up the Fact module alongside the existing memory system; wrap writes with a datom-emitting interceptor. Zero reads change.
\item \textbf{Backfill (3 days).} Synthesize datoms from existing mutable state with \C{tx-time = created\_at}, \C{valid-from = created\_at}.
\item \textbf{Query surface (1 week).} Build \C{as-of}, \C{history}, \C{branch} as thin functions over the log; existing skills rewire to use them.
\end{enumerate}

Subsequent migration of Effect, Spec, and Replay (Phase 1) proceeds one module at a time, at the operator's pace. Plan, Txn, and REPL land in Phase 2.

\subsection{Relation to neurosymbolic AI}\label{sec:nesy-rel}

Persistence is neurosymbolic in a specific technical sense: the \emph{substrate} is symbolic and the \emph{agency} is neural. Phase 1 ships four symbolic substrate pieces:

\begin{itemize}
\item \textbf{Bitemporal datom queries with a Datalog-shaped surface} --- \C{as\_of}, \C{history}, \C{branch}, \C{validAsOf} are relational queries over an $(e, a, v, \tau, \dots)$ tuple store; Phase 1 implements these as Python comprehensions over the log (a Datalog query helper is adjacent future work --- see below).
\item \textbf{EDN AST grammars for plans} --- \C{:persistence.plan/node} is registered in Phase 1; the Plan-module evaluator is Phase 2.
\item \textbf{Policy-as-data} --- shipped (\C{policy\_eval.py}, \textasciitilde{}200 LOC).
\item \textbf{Malli-style specs with LLM self-healing hints} --- shipped (\C{persistence.spec}, 186 tests; \C{explain\_for\_llm}, \sectionref{sec:specs}).
\end{itemize}

Two items we explicitly do \emph{not} claim as Phase 1 shipped are Datalog and Z3. We separate them cleanly from shipped capability:

\begin{itemize}
\item \textbf{Datalog engine.} The datom model is Datalog-ready --- classical Datalog rules over $(e, a, v)$ triples sit naturally on top of our query surface --- but the Phase 1 query layer is Python list comprehensions, not a rule engine. A Datalog surface (a compact pattern-matching helper over the datom set) is an adjacent system we draw on and is straightforward Phase 2 work.
\item \textbf{Z3-discharged \C{verify} leaves.} Proof-of-Thought~\citep{fan2024proofofthought} describes a compelling integration point for Z3 at \C{verify}-node leaves, where plan fragments discharge proof obligations against an SMT solver. This is a future Plan-module feature; no Z3 code ships in Phase 1.
\end{itemize}

The symbolic substrate we \emph{do} ship --- bitemporal log, handler stack with \C{is\_well\_formed} / \C{Unhandled}, Merkle-hashed audit chain, boundary specs with \C{explain\_for\_llm} --- carries the properties we care about: accountability, composition, formal guarantees, and a concrete LLM-self-healing contract. The neural layer (LLM decisions, embeddings, generative skills) carries expressiveness. This distinguishes Persistence from pure-neural approaches (no guarantees) and pure-symbolic approaches (no expressiveness), and positions it as a candidate substrate for neurosymbolic systems beyond LLM agents, including embodied agents, multi-agent planning, and human-in-the-loop scientific discovery.

\section{Conclusion}\label{sec:conclusion}

We presented Persistence, a cognitive-runtime substrate for accountable neurosymbolic agents. Phase 1 plus the first Phase-2 increment of the reference implementation (v0.1.0a1 + v0.4.0a1, \textbf{832 tests green + 7 xfailed}) ships five modules --- Fact, Effect, Spec, Replay, Plan (minimal) --- demonstrating five substrate-derived capabilities: queryable bitemporal history, counterfactual branching with parent-store isolation, composable policy-gated effects with a machine-checkable well-formedness property, boundary-checked contracts with LLM-self-healing hints, and content-addressed homoiconic plan ASTs with byte-identical round-trip and descendant-propagation identity. Five formal propositions hold on the shipped artifact; the headline two are: well-formedness of any handler stack is decidable in linear time by \C{Runtime.is\_well\_formed} and enforced at runtime by \C{Unhandled}; replay of a NO-OP intervention produces a trajectory whose canonical hash is byte-identical to the factual trajectory's. The Merkle-hashed audit chain (\C{verify\_chain}) shipped and is the regulator-grade foundation on which Case A (project finance) and Case C (insurance) will build in Phase 2. Three further modules (Plan execution, Txn, REPL) are specified, partially registered in the spec registry ahead of the runtime (a deliberate parse-don't-validate methodology), and scheduled for Phase 2 (2026-Q3).

The core claim is that accountability, counterfactual replay, composable safety, compositional skill learning, multi-agent coordination, and live production steering can be derived from one substrate rather than five --- and that the four already shipped are \emph{properties of the log, checkable against the artifact}, not features engineered per agent.

\textbf{Future work.} We plan to extend the runtime to (1) a persistent-trie \C{Store} backend that makes Proposition~\ref{prop:branch}'s $O(\log n)$ branch share a shipped property, not an idealized-implementation theorem; (2) per-step rng-state recording in the replay engine, which unblocks byte-identical replay for LLM trajectories; (3) ed25519 per-transaction signing via a \C{sign\_transaction} primitive verified in \C{verify\_chain}; (4) a Datalog query surface over the datom set; (5) Z3-discharged \C{verify} leaves in the Plan-module AST following Proof-of-Thought~\citep{fan2024proofofthought}; (6) embodied-agent extensions where effects include physical-world interactions with partial observability; and (7) distributed deployment preserving bitemporal semantics across multi-region fact stores.

% RESTORE: 'We release the Phase-1 runtime under AGPL-3, with paper artifacts (benchmark harness, 50-trajectory synthetic regulator-replay dataset) under CC-BY-4.0.'

% Acknowledgements — \acks{} is JMLR-native; auto-suppressed under [anon].
\acks{% Set under [final] before camera-ready.
}

\bibliography{references}   % no \bibliographystyle — class controls it

\end{document}
